Drought frequency predicts life history strategies in \emph{Heliophila}

### 2018-12-10 manuscript/manuscript.tex
remove firth

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons.
These represent fundamentally different life history strategies, but the
ecological factors that explain their evolution and distributions remain
empirically uresolved (Friedman and Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov and Schaffer, 1973; Franco and Silvertown,
1996; Stearns, 1992). In plants, this has been extended to the
hypothesis that annuality is adaptive when it allows plants to escape
drought (Schaffer and Gadgil, 1975). Lack of water is perhaps the
greatest threat to survival during vegetative or reproductive growth and
annuals can remain dormant (and protected as a seed) during drought.
Thus, environments with greater seasonal drought frequency may select
for annual life histories that complete reproduction prior to drought
prone seasons. Conversely, environments with less frequent drought may
select for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin and D'Antonio, 2004). These predictions have been
supported by the observation of annuals in arid environments in
\emph{Oryza perennis} (Morishima et al., 1984) and \emph{Oenothera}
(Evans et al., 2005). Additionally, annual and perennial species of
\emph{Nemesia} were qualitatively associated with winter rather and
summer rainfall environments respectively (Datson et al., 2008) and
annual species of \emph{Scorzoneroides} were associated with
environments classified as unpredictable (Cruz-Mazo et al., 2009).
However, whether the history frequency of drought events indeed predicts
the distributions annual or perennial life history strategies has yet to
be tested.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses within the African endemic mustard genus,
\emph{Heliophila} L. (Brassicaceae). If annuality is an adaptive
strategy allowing plants to escape drought prone seasons, then drought
frequency should predict the distribution of life history strategies
across landscapes, and annual species should be more commonly associated
with drought prone regions than perennial species. Furthermore, if
annual species have adapted to escape drought prone seasons,
observations of growing annual species (i.e.~occurring in forms other
than seed) should be rare during drought prone seasons. Phylogenetic
relatedness can have significant non-random effects on species
distributions and life history traits (Barrett et al., 1996), and
therefore we assessed the relationship between life history distribution
and drought frequency in a phylogenetically controlled background.

@@

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas et al., 2011; Kogan et al.,
2016). But they also present a new tool to study seasonal patterns in
the frequency of drought across environments and to test hypotheses
about the effect of drought on ecological and evolutionary processes
(Kerr and Ostrovsky, 2003). As such, the VHI has been applied recently
to study drought related ecology of natural species and proven useful
for predicting intraspecific variation in drought tolerance traits and
genes (Dittberner et al., 2018; Mojica et al., 2016; John G Monroe et
al., 2018). Here, we accessed VHI data at \(16km^2\) resolution from
1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species.

\hypertarget{life-history-data-for_-heliophila_}{%
\subsubsection{Life history data for\_
Heliophila\_}\label{life-history-data-for_-heliophila_}}

@@

\#\#\#\_ Heliophila\_ occurrence records

@@

Aligned\_ Heliophila\_ ITS sequences were obtained from previous work by
Mandáková et al. (2012). Aethionema, Alliaria, Cardamine, Chamira, and
Rorippa ITS records from were downloaded from Genbank.

@@

Out group species were aligned together with \emph{Heliophila} ITS
sequences using MAFFT. Model selection for construction of phylogeny was
performed in jModeltest2 with CIPRES. Based on this analysis,
\(GTR + L\) were selected. Ultrametric phylogeny was estimated with
branch lengths as relative time.

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought during the winter, spring, summer, and fall between
raw occurrence records of annual and perennial species by t-tests. To
account for variation in the number of occurrence records per species,
we next calculated the mean drought frequency during the winter, spring,
summer and fall for each species. The relationships between species mean
values of drought frequency during each season and life habitat (annual
or perennial) were tested using Firth's penalized-likelihood logistic
regressions. Because demographic histories caused by ancestry can
confound trait - environment associations, we then tested for the
relationships between drought frequency and life history while
controlling for relationships between species using phylogenetic
logistic regressions.

@@

Theory predicts that annuality should be adaptive in places where
stresses such as drought are more common. Conversely, perenniality
should be adaptive in places where such stresses are less frequent. We
found that the frequency of drought was significantly higher at the
locations of occurrence records for annual species. In terms of raw
observation records (Figure \ref{fig:mapsboxplots}g), the frequency of
drought was significantly higher at the location of annuals during the
winter (t = 10.65, p = 0.00), spring (t = 10.73, p = 0.00), summer (t =
12.67, p = 0.00), and fall (t = 15.26, p = 0.00). Because raw occurrence
records do not account for variation in the number of records per
species (Table \ref{tab:speciesmeanstable}), we also tested whether mean
drought frequency values of each species were significantly different
between annuals and perennials by a Firths-penalized logistic
regression. We found that the mean drought frequencies were
significantly higher (\(\alpha = 0.05\)) in annual species during the
spring, summer, and fall (Table \ref{tab:modelstable}, Figure
\ref{fig:lineplots}a). We further tested whether annual species are
found in places where droughts occur more frequently while controlling
for the phylogentic relatedness of \emph{Heliophila} species. This is
important, because environmental differences in species distributions
can be confounded with demographic history caused by ancestry. We found
that while controlling for phylogeny, the mean drought frequencies were
significantly higher (\(\alpha = 0.05\)) in annual species during the
spring, summer, and fall (Table \ref{tab:modelstable}, Figure
\ref{fig:lineplots}a). These findings indicate that common acestry alone
does not explain differences the drought frequencies experienced in the
environments of annual and perennial \emph{Heliophila}.

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:modelstable}Logistic regressions between life history, and the mean drought frequency observed at herbaria collection sites of\_ Heliophila\_ species the winter, spring, summer, and fall.}
\begin{tabular}{lllll}
\toprule
Predictor & \multicolumn{1}{c}{Estimate`} & \multicolumn{1}{c}{P`} & \multicolumn{1}{c}{Estimate*} & \multicolumn{1}{c}{P*}\\
\midrule
Intercept & 2.2575 & 0.1739 & 0.7231 & 0.6636\\
Winter drought freq. & -6.7484 & 0.1661 & -1.5452 & 0.7274\\ \midrule
Intercept & 4.5594 & 0.0443 & 5.0107 & 0.0534\\
Spring drought freq. & -11.7895 & 0.0423 & -12.9014 & 0.0464\\ \midrule
Intercept & 7.1742 & 0.0011 & 7.7093 & 0.0054\\
Summer drought freq. & -18.2999 & 0.0010 & -19.9056 & 0.0042\\ \midrule
Intercept & 6.4226 & 0.0029 & 7.0162 & 0.0082\\
Fall drought freq. & -19.0512 & 0.0026 & -20.8174 & 0.0067\\ \midrule
\bottomrule
\addlinespace
\end{tabular}
\begin{tablenotes}[para]
\normalsize{\textit{Note.} ` = Firth's penalized logistic regression. * = Phylogenetically constrained logistic regression. Annual species were scored as 0 and perennial species as 1.}
\end{tablenotes}
\end{threeparttable}
\end{center}
\end{table}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/line_and_dates} \caption{(a) Comparison (mean +- SE) of drought frequency across
seasons measured at the GBIF records of annual and perennial species
of\_ Heliophila\_. (phylogenetic logistic regression, * = p \textless{}
0.05, ** = p \textless{} 0.01) (b) Collection dates of GBIF records of
annual and perennial species of\_ Heliophila\_.}\label{fig:lineplots}
\end{figure}

@@

To test the hypothesis that annual and perennial plants reflect
adaptation to alternative drought environments we examined the landscape
distribution of life history strategies in the large and diverse mustard
genus, \emph{Heliophila}. Using metadata of 2192 occurrence records and
a 34 year dataset of satellite-detected droughts, we tested the
prediction that annual species are more often observed in drought-prone
locations than perennial species, when controlling for phylogenetic
relatedness. We found that drought frequency is significantly different
between the distributions of annual and perennial species, with annuals
being found in environments with more frequent drought, and that this
signal is strongest during the seasons when annuals are likely escaping
via seed dormancy These results remain significant while controlling for
the phylogenetic relationships of \emph{Heliophila} species, yielding
support for the role that natural selection has played in driving
contemporary distributions of these alternatives strategies in relation
to drought regimens.

We cannot eliminate the possibility that confounding traits or
environmental variables are the causative factors explaining variation
in the distributions of annual and perennial species. Nevertheless,
these results reveal quantitative support for the classic prediction
that annual species are found in environemnts that experience more
frequent drought than perennial species. These findings complement
previous reports of qualitative associations between annuality with
enviroments characterized as having increased aridity (Evans et al.,
2005), alternative precipitation defined habitats (Datson et al., 2008;
Morishima et al., 1984), or greater unpredictability (Cruz-Mazo et al.,
2009). However, to our knowledge this is the first study to demonstrate
a significant association between life history and drought in a
phylogenetic context informed by large scale species distribution data
and long term drought measures.

@@

These findings support classical theoretical predictions about the
adaptive value of annual and perennial life history strategies. Taken
together, they suggest that in \emph{Heliophila}, annual species are
adapted to environments with increased summer droughts by avoiding these
seasons in a dormant seed phase of their life cycle. They also suggest
that perenniality is adaptive in environemnts where droughts are less
frequent. While most previous work has focused on describing the
evolutionary origins of annuality (Andreasen and Baldwin, 2001; Barrett
et al., 1996; Conti et al., 1999; Friedman and Rubin, 2015; Verboom et
al., 2004) there are at least a few other cases where perenniality
appears to have arisen from an annual ancestor (Bena et al., 1998; Tank
and Olmstead, 2008). And while early theory predicted selection for
annuality when adult morality is high (Stearns, 1992), we also find
evidence that the transition to perenniality could be explained by the
historical frequency of drought, a likely cause of mortality in plants.
The phylogeny reveals several transitions from annual to perennial life
history (Figure \ref{fig:phylogeny}a) and that the distributions of
perennial \emph{Heliophila} extend into regions where drought frequency
is low (Figure \ref{fig:mapsboxplots}b,Figure \ref{fig:speciesmaps}).
Perennials may be able to out complete annual relatives in environments
where the infrequency of drought favors strategies that allow plants to
benefit from growth over many seasons. It may also indicate that annuals
rely on drought as a source of disturbance for seedling recruitment when
competing with perennials (Corbin and D'Antonio, 2004). Indeed, no
annual species were observed in the low drought regions of eastern South
Africa \ref{fig:mapsboxplots}a,Figure \ref{fig:speciesmaps}).

These findings suggest that species with locally adaptive life history
strategies could be threatened by rapidly changing drought regimens
(Dai, 2011). This could have impacts on ecosystem functioning and
processes such as carbon cycling if the composition of annual and
perennial species changes as a response (Garnier et al., 1997; J Grey
Monroe et al., 2018; Roumet et al., 2006). Furthermore, the frequency of
drought may be an important factor when considering the use of perennial
cropping systems (Lelièvre and Volaire, 2009; Parry et al., 2005).

@@

@@

@@

@@

@@

@@

@@

\leavevmode\hypertarget{ref-monroe2018ecoevolutionary}{}%
Monroe JG, Markman DW, Beck WS, Felton AJ, Vahsen ML, Pressler Y. 2018.
Ecoevolutionary dynamics of carbon cycling in the anthropocene.
\emph{Trends in ecology \& evolution} \textbf{33}:213--225.

\leavevmode\hypertarget{ref-monroe2018drought}{}%
Monroe JG, Powell T, Price N, Mullen J, Howard A, Evans K, Lovell J,
McKay J. 2018. Drought adaptation in nature by extensive genetic
loss-of-function. \emph{bioRxiv} 372854.

@@

We used R (Version 3.5.1; R Core Team, 2018) and the R-packages
\emph{ape} (Version 5.2; Paradis and Schliep, 2018; Orme et al., 2018;
Soetaert, 2018), \emph{bindrcpp} (Version 0.2.2; Müller, 2018),
\emph{caper} (Version 1.0.1; Orme et al., 2018), \emph{coda} (Version
0.19.2; Plummer et al., 2006), \emph{diagram} (Version 1.6.4; Soetaert,
2017), \emph{dplyr} (Version 0.7.8; Wickham et al., 2018),
\emph{forcats} (Version 0.3.0; Wickham, 2018a), \emph{gee} (Version
4.13.19; R by Thomas Lumley and author., 2015), \emph{geiger} (Version
2.0.6; Alfaro et al., 2009; Eastman et al., 2011; Harmon et al., 2008;
Slater et al., 2012), \emph{ggplot2} (Version 3.1.0; Wickham, 2016),
\emph{logistf} (Version 1.23; Heinze and Ploner, 2018), \emph{maps}
(Version 3.3.0; Richard A. Becker et al., 2018), \emph{MASS} (Version
7.3.51.1; Venables and Ripley, 2002), \emph{Matrix} (Version 1.2.15;
Bates and Maechler, 2018), \emph{MCMCglmm} (Version 2.26; Hadfield,
2010), \emph{mvtnorm} (Version 1.0.8; Genz and Bretz, 2009),
\emph{papaja} (Version 0.1.0.9842; Aust and Barth, 2018), \emph{phylolm}
(Version 2.6; Ho and Ane, 2014), \emph{phytools} (Version 0.6.60;
Revell, 2012), \emph{purrr} (Version 0.2.5; Henry and Wickham, 2018),
\emph{raster} (Version 2.8.4; Hijmans, 2018), \emph{readr} (Version
1.2.1; Wickham et al., 2017), \emph{shape} (Version 1.4.4; Soetaert,
2018), \emph{sp} (Version 1.3.1; Pebesma and Bivand, 2005),
\emph{stringr} (Version 1.3.1; Wickham, 2018b), \emph{tibble} (Version
1.4.2; Müller and Wickham, 2018), \emph{tidyr} (Version 0.8.2; Wickham
and Henry, 2018), and \emph{tidyverse} (Version 1.2.1; Wickham, 2017)
for all our analyses.

>>>>

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons.
These represent fundamentally different life history strategies, but the
ecological factors that explain their evolution and distributions remain
empirically uresolved (Friedman and Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov and Schaffer, 1973; Franco and Silvertown,
1996; Stearns, 1992). In plants, this has been extended to the
hypothesis that annuality is adaptive when it allows plants to escape
drought (Schaffer and Gadgil, 1975). Lack of water is perhaps the
greatest threat to survival during vegetative or reproductive growth and
annuals can remain dormant (and protected as a seed) during drought.
Thus, environments with greater seasonal drought frequency may select
for annual life histories that complete reproduction prior to drought
prone seasons. Conversely, environments with less frequent drought may
select for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin and D'Antonio, 2004). These predictions have been
supported by the observation of annuals in arid environments in
\emph{Oryza perennis} (Morishima et al., 1984) and \emph{Oenothera}
(Evans et al., 2005). Additionally, annual and perennial species of
\emph{Nemesia} were qualitatively associated with winter rather and
summer rainfall environments respectively (Datson et al., 2008) and
annual species of \emph{Scorzoneroides} were associated with
environments classified as unpredictable (Cruz-Mazo et al., 2009).
However, whether the history frequency of drought events indeed predicts
the distributions annual or perennial life history strategies has yet to
be tested.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses within the African endemic mustard genus,
\emph{Heliophila} L. (Brassicaceae). If annuality is an adaptive
strategy allowing plants to escape drought prone seasons, then drought
frequency should predict the distribution of life history strategies
across landscapes, and annual species should be more commonly associated
with drought prone regions than perennial species. Furthermore, if
annual species have adapted to escape drought prone seasons,
observations of growing annual species (i.e.~occurring in forms other
than seed) should be rare during drought prone seasons. Phylogenetic
relatedness can influence tests of associations between species' traits
and their environments (Barrett et al., 1996; Felsenstein, 1985), and
therefore we assessed the relationship between life history distribution
and drought frequency in a phylogenetic context.

@@

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas et al., 2011; Kogan et al.,
2016). But they also present a new tool to study seasonal patterns in
the frequency of drought across environments and to test hypotheses
about the effect of drought on ecological and evolutionary processes
(Kerr and Ostrovsky, 2003). As such, the VHI has been applied recently
to study drought related ecology of natural species and proven useful
for predicting intraspecific variation in drought tolerance traits and
genes (Dittberner et al., 2018; Mojica et al., 2016; J. Monroe, Powell,
et al., 2018). Here, we accessed VHI data at \(16km^2\) resolution from
1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species.

\hypertarget{life-history-data-for-heliophila}{%
\subsubsection{\texorpdfstring{Life history data for
\emph{Heliophila}}{Life history data for Heliophila}}\label{life-history-data-for-heliophila}}

@@

\hypertarget{heliophila-occurrence-records}{%
\subsubsection{\texorpdfstring{\emph{Heliophila} occurrence
records}{Heliophila occurrence records}}\label{heliophila-occurrence-records}}

@@

An alignment of ITS I and II sequences for \emph{Heliophila} species was
obtained from the authors of Mandáková et al. (2012). Individual ITS I
and II sequences for \emph{Aethionema grandiflorum, Alliaria petiolata,
Cardamine matthioli, Chamira circaeoides}, and \emph{Rorippa amphibia}
were downloaded from Genbank.

@@

Out group ( \emph{Aethionema grandiflorum, Alliaria petiolata, Cardamine
matthioli, Chamira circaeoides}, and \emph{Rorippa amphibia}) and
ingroup \emph{Heliophila} ITS I and II sequences were aaligned using
MAFFT (Katoh et al., 2002) with strategy G-INS-I, offset value 0.1, and
all other options set as default. The \(GTR + L\) model of nucleotide
substitution was determined to best fit the data based on AIC using
jModelTest2 (Darriba et al., 2012; Guindon and Gascuel, 2003). A maximum
clade credibility tree with branch lengths as relative time was
estimated by summarizing data from six runs of 100,000,000 generations
of Bayesian Markov chain Monte Carlo conducted in BEAST 2 (Bouckaert et
al., 2014). Model selection and phylogenetic analyses were conducted
through the CIPRES Science Gateway (Miller et al., 2010).

@@

To evaluate the hypothesis that annual and perennial life history
strategies reflect adaptations to alternative drought regimes, we tested
the corresponding prediction that the observed distributions of annual
and perennial \emph{Heliophila} species would be significantly
associated with historic drought frequency. First, we compared the
frequency of drought during the winter, spring, summer, and fall between
raw occurrence records of annual and perennial species by t-tests. To
account for variation in the number of occurrence records per species,
we next calculated the mean drought frequency during the winter, spring,
summer and fall for each species. Because shared evolutionary history of
closely related species can lead to spurious associations between traits
and environments (Felsenstein, 1985), we tested for a relationship
between life history strategy and drought frequency while controlling
for phylogeny using phylogenetic logistic regression (Ives and Garland,
2010).

@@

Theory predicts that annuality should be adaptive in places where
stresses such as drought are more common. Conversely, perenniality
should be adaptive in places where such stresses are less frequent. We
found that the frequency of drought was significantly higher at the
locations of occurrence records for annual species. In terms of raw
observation records (Figure \ref{fig:mapsboxplots}g), the frequency of
drought was significantly higher at the location of annuals during the
winter (t = 10.65, p = 0.00), spring (t = 10.73, p = 0.00), summer (t =
12.67, p = 0.00), and fall (t = 15.26, p = 0.00). Because raw occurrence
records do not account for variation in the number of records per
species (Table \ref{tab:speciesmeanstable}) or species relatedness
(Figure \ref{fig:phylogeny}a), we also tested whether mean drought
frequency values of each species were significantly different between
annuals and perennials using phylogenetic logistic regression. We found
that the mean drought frequencies were significantly higher
(\(\alpha = 0.05\)) in annual species during the spring, summer, and
fall (Table \ref{tab:modelstable}, Figure \ref{fig:lineplots}a). These
findings indicate that common acestry alone does not explain differences
the drought frequencies experienced in the environments of annual and
perennial \emph{Heliophila}.

\begin{table}[tbp]
\begin{center}
\begin{threeparttable}
\caption{\label{tab:modelstable}Phylogenetic logistic regressions between life history, and the mean drought frequency observed at herbaria collection sites of Heliophila species the winter, spring, summer, and fall.}
\begin{tabular}{lll}
\toprule
Predictor & \multicolumn{1}{c}{Estimate} & \multicolumn{1}{c}{P}\\
\midrule
Intercept & 0.7231 & 0.6636\\
Winter drought freq. & -1.5452 & 0.7274\\ \midrule
Intercept & 5.0107 & 0.0534\\
Spring drought freq. & -12.9014 & 0.0464\\ \midrule
Intercept & 7.7093 & 0.0054\\
Summer drought freq. & -19.9056 & 0.0042\\ \midrule
Intercept & 7.0162 & 0.0082\\
Fall drought freq. & -20.8174 & 0.0067\\ \midrule
\bottomrule
\addlinespace
\end{tabular}
\begin{tablenotes}[para]
\normalsize{\textit{Note.} Annual species were scored as 0 and perennial species as 1.}
\end{tablenotes}
\end{threeparttable}
\end{center}
\end{table}

\begin{figure}[!h]
\includegraphics[width=\textwidth]{../figures/line_and_dates} \caption{(a) Comparison (mean +- SE) of drought frequency across
seasons measured at the GBIF records of annual and perennial species of
\emph{Heliophila}. (phylogenetic logistic regression, * = p \textless{}
0.05, ** = p \textless{} 0.01) (b) Collection dates of GBIF records of
annual and perennial species of \emph{Heliophila}.}\label{fig:lineplots}
\end{figure}

@@

To test the hypothesis that annual and perennial plants reflect
adaptation to alternative drought environments we examined the landscape
distribution of life history strategies in the large and diverse mustard
genus, \emph{Heliophila}. Using metadata of 2192 occurrence records and
a 34 year dataset of satellite-detected droughts, we tested the
prediction that annual species are more often observed in drought-prone
locations than perennial species, when controlling for phylogenetic
relatedness. We found that drought frequency is significantly different
between the distributions of annual and perennial species, with annuals
being found in environments with more frequent drought, and that this
signal is strongest during the seasons when annuals are likely escaping
via seed dormancy. These results remain significant while controlling
for the phylogenetic relationships of \emph{Heliophila} species,
yielding support for the role that natural selection has played in
driving contemporary distributions of these alternatives strategies in
relation to drought regimens.

We cannot eliminate the possibility that confounding traits or
environmental variables are the causative factors explaining variation
in the distributions of annual and perennial species. Nevertheless,
these results provide quantitative support for the classic prediction
that annual species are found in environemnts that experience more
frequent drought than perennial species. These findings complement
previous reports of qualitative associations between annuality with
enviroments characterized as having increased aridity (Evans et al.,
2005), alternative precipitation defined habitats (Datson et al., 2008;
Morishima et al., 1984), or greater unpredictability (Cruz-Mazo et al.,
2009). However, to our knowledge this is the first study to demonstrate
a significant association between life history and drought in a
phylogenetic context informed by large scale species distribution data
and long term drought measures.

@@

These findings support classical theoretical predictions about the
adaptive value of annual and perennial life history strategies. Taken
together, they suggest that in \emph{Heliophila}, annual species are
adapted to environments with increased summer droughts by avoiding these
seasons in a dormant seed phase of their life cycle. They also suggest
that perenniality is adaptive in environemnts where droughts are less
frequent. While most previous work has focused on describing the
evolutionary origins of annuality (Andreasen and Baldwin, 2001; Barrett
et al., 1996; Conti et al., 1999; Friedman and Rubin, 2015; Verboom et
al., 2004) there are at least a few other cases where perenniality
appears to have arisen from an annual ancestor (Bena et al., 1998; Tank
and Olmstead, 2008). And while early theory predicted selection for
annuality when adult morality is high (Stearns, 1992), we also find
evidence that the transition to perenniality could be explained by the
historical frequency of drought, a likely cause of mortality in plants.
The phylogeny reveals several transitions from annual to perennial life
history (Figure \ref{fig:phylogeny}a) and that the distributions of
perennial \emph{Heliophila} extend into regions where drought frequency
is low (Figure \ref{fig:mapsboxplots}b, Figure \ref{fig:speciesmaps}).
Perennials may be able to out complete annual relatives in environments
where the infrequency of drought favors strategies that allow plants to
benefit from growth over many seasons. It may also indicate that annuals
rely on drought as a source of disturbance for seedling recruitment when
competing with perennials (Corbin and D'Antonio, 2004). Indeed, no
annual species were observed in the low drought regions of eastern South
Africa (Figure \ref{fig:mapsboxplots}, Figure \ref{fig:speciesmaps}).

These findings suggest that species with locally adaptive life history
strategies could be threatened by rapidly changing drought regimens
(Dai, 2011). This could have impacts on ecosystem functioning and
processes such as carbon cycling if life history traits evolve or the
composition of annual and perennial species changes in response (Garnier
et al., 1997; J. Monroe, Markman, et al., 2018; Roumet et al., 2006).
Furthermore, the frequency of drought may be an important factor when
considering the use of perennial cropping systems (Lelièvre and Volaire,
2009; Parry et al., 2005).

@@

\leavevmode\hypertarget{ref-bouckaert2014beast}{}%
Bouckaert R, Heled J, Kühnert D, Vaughan T, Wu C-H, Xie D, Suchard M,
Rambaut A, Drummond A. 2014. BEAST 2: A software platform for bayesian
evolutionary analysis. \emph{PLoS Computational Biology}
\textbf{10}:doi:10.1371/journal.pcbi.1003537.

@@

\leavevmode\hypertarget{ref-darriba2012jmodeltest}{}%
Darriba D, Taboada G, Doallo R, Posada D. 2012. JModelTest 2: More
models, new heuristics and parallel computing. \emph{Nature Methods}
\textbf{9}:772.

@@

\leavevmode\hypertarget{ref-felsenstein1985phylogenies}{}%
Felsenstein J. 1985. Phylogenies and the comparative method.
\emph{American Naturalist} \textbf{125}:1--15.

@@

\leavevmode\hypertarget{ref-guindon2003a}{}%
Guindon S, Gascuel O. 2003. A simple, fast and accurate method to
estimate large phylogenies by maximum-likelihood. \emph{Systematic
Biology} \textbf{52}:696--704.

@@

\leavevmode\hypertarget{ref-ives2010phylogenetic}{}%
Ives A, Garland T. 2010. CPhylogenetic logistic regression for binary
dependent variables. \emph{Systematic Biology} \textbf{59}:9--26.

\leavevmode\hypertarget{ref-katoh2002mafft}{}%
Katoh, Misawa, Kuma, Miyata. 2002. MAFFT: A novel method for rapid
multiple sequence alignment based on fast fourier transform.
\emph{Nucleic Acids Research} \textbf{30}:3059--3066.

@@

\leavevmode\hypertarget{ref-miller2010creating}{}%
Miller M, Pfeiffer W, Schwartz T and. 2010. Creating the cipres science
gateway for inference of large phylogenetic trees. \emph{Proceedings of
the Gateway Computing Environments Workshop} 1--8.

@@

\leavevmode\hypertarget{ref-monroe2018ecoevolutionary}{}%
Monroe J, Markman D, Beck W, Felton A, Vahsen M, Pressler Y. 2018.
Ecoevolutionary dynamics of carbon cycling in the anthropocene.
\emph{Trends in ecology \& evolution} \textbf{33}:213--225.

\leavevmode\hypertarget{ref-monroe2018drought}{}%
Monroe J, Powell T, Price N, Mullen J, Howard A, Evans K, Lovell J,
McKay J. 2018. Drought adaptation in nature by extensive genetic
loss-of-function. \emph{eLife} DOI: 10.7554/eLife.41038.

@@

We used R (Version 3.5.1; R Core Team, 2018) and the R-packages
\emph{ape} (Version 5.2; Paradis and Schliep, 2018; Orme et al., 2018;
Soetaert, 2018), \emph{bindrcpp} (Version 0.2.2; Müller, 2018),
\emph{caper} (Version 1.0.1; Orme et al., 2018), \emph{coda} (Version
0.19.2; Plummer et al., 2006), \emph{diagram} (Version 1.6.4; Soetaert,
2017), \emph{dplyr} (Version 0.7.8; Wickham et al., 2018),
\emph{forcats} (Version 0.3.0; Wickham, 2018a), \emph{gee} (Version
4.13.19; R by Thomas Lumley and author., 2015), \emph{geiger} (Version
2.0.6; Alfaro et al., 2009; Eastman et al., 2011; Harmon et al., 2008;
Slater et al., 2012), \emph{ggplot2} (Version 3.1.0; Wickham, 2016),
\emph{logistf} (Version 1.23; Heinze and Ploner, 2018), \emph{maps}
(Version 3.3.0; Richard A. Becker et al., 2018), \emph{MASS} (Version
7.3.51.1; Venables and Ripley, 2002), \emph{Matrix} (Version 1.2.15;
Bates and Maechler, 2018), \emph{MCMCglmm} (Version 2.26; Hadfield,
2010), \emph{mvtnorm} (Version 1.0.8; Genz and Bretz, 2009),
\emph{papaja} (Version 0.1.0.9842; Aust and Barth, 2018), \emph{phylolm}
(Version 2.6; Ho and Ane, 2014), \emph{phytools} (Version 0.6.60;
Revell, 2012), \emph{purrr} (Version 0.2.5; Henry and Wickham, 2018),
\emph{raster} (Version 2.8.4; Hijmans, 2018), \emph{readr} (Version
1.2.1; Wickham et al., 2017), \emph{shape} (Version 1.4.4; Soetaert,
2018), \emph{sp} (Version 1.3.1; Pebesma and Bivand, 2005),
\emph{stringr} (Version 1.3.1; Wickham, 2018b), \emph{tibble} (Version
1.4.2; Müller and Wickham, 2018), \emph{tidyr} (Version 0.8.2; Wickham
and Henry, 2018), and \emph{tidyverse} (Version 1.2.1; Wickham, 2017)
for all our analyses. \newpage
